---
title: "Pandas HDF5 to R"
author: "Martin Ueding"
date: "27 November 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Motivation

Markus' subduction code is written in Python using the Pandas library. Said library brings the “data frame” concept into Python; an additional feature is the hierarchical multi index (*HMI* from here on). Pandas can store data to HDF5, but it does this in a seemingly strange way. This format cannot readily be imported into R-hadron, therefore we need to massage it into the right structure first. This notebook will contain a function to do the needed work.

# Exploring the data

```{r}
library(rhdf5)
```

A sample file produced with the subduction code has the following structure:

```{r}
filename <- 'C20_p1_A1.h5'
knitr::kable(h5ls('C20_p1_A1.h5'))
```

There are a couple of concepts:

`axis`
:   We store 2D data, therefore we have rows and columns. `axis0` describes the columns, though this seems unusual. `axis1` describes the rows.

`label` and `level`
:   A “factor” in statistics is discrete variable, it can be nominal or ordinal, the latter having an ordering. In the process of “coding” one assigns a certain value an index. Say you have “male” and “female” and code this with “0” and “1” in your data. We would call $\{ \mathrm{male}, \mathrm{female} \}$ the “levels” and ${0, 1}$ the unique “labels”, because that is what you assigned to the levels. The data then is represented by the labels only, because we can always just get the levels back.

:   Pandas stores the row and column names as such factors, even if they are just the numbers from 1 to $N$. The number of labels in a factor matches the number of rows (or column) that this factor belongs to.

`block_items`
:   We haven't figured that out yet. It seems to be redundant for our needs.

`block_values`
:   The actual matrix of values, so in R language this would be `as.matrix(df)` for some `data.frame` called `df`.

There are a couple of lengths here:

33
:   The number of combinations of indices that we have. This number depends on the number irreps, $\Gamma$ structures and the like.

32
:   Number of configurations. It is a bit unfortunate, but with a bit of care we should be able to figure this out.

48
:   Halved time extent of the lattice

1536
:   Since $32 \cdot 48 = 1536$, this is the number of result values (time slice × configuration).

R-hadron uses a 2D array for storing the “time slice × configuration” array, so we want to do the same thing here.

# Implementation

## Step by step

We will first start to lay out the individual pieces and then wrap it up in one convenient function which does all the magic.

### Loading data

Loading the data gives us a list with the names (“dict” with “keys” in Python nomenclature) that we would expect:

```{r}
data <- h5read(filename, 'data')
names(data)
```

These warnings do not look good, we will have to check whether they are a problem later on.

### Payload

Basically we now want to create a data frame which has 2D arrays as values. Then it has a bunch of columns which have all the `axis1` indices as columns. We want to use the “long format” here, so there is no fancy HMI, it will be a repetition that later on has to be worked with the `dplyr` package, which is really great at that.

The `block0_values` contains real and imaginary part. They are separated, therefore we first combine them into one block.

```{r}
block <- data$block0_values$r + 1i * data$block0_values$i
```

We are also now rid of the additional hierarchy and can mold this into a list of matrices. The version with `apply(block, 1, ...)` did not allow us to return a list, so instead we do the following with indices.

```{r}
# Lattice time extent
t <- 48
block.list <- lapply(1:ncol(block), function (i) t(matrix(block[ , i], nrow = t)))
```

Now we have a list with matrices. Perhaps one has to transpose the data structure in order to get the right row-major or column-major layout.

### Meta data

From here on we just need to convert the `axis1` elements into factors and add them to the `block.list` data structure. Let's take the 8th axis as an example. Its levels (the values that it can possible take) are these:

```{r}
data$axis1_level8
```

The *indices* of the levels that are taken by the 33 data points are the following:

```{r}
data$axis1_label8
```

This list is 0-indexed, R uses 1-indexing. We want to “inflate” this relation now and then let R “deflate” it using the `factor` again:

```{r}
key8 <- factor(x = data$axis1_level8[data$axis1_label8 + 1], levels = data$axis1_level8)
key8
```

That's exactly what we want. A 33 element long thing that we can use as a column in our resulting data frame.

We would ideally like to convert all meta data automatically, no matter how many there are.

```{r}
meta.cols <- sum(startsWith(names(data), 'axis1_label'))
```


### Test assembly

We can try to assemble a data frame with this before we cast it into a function. [Hadley Wickam notes](https://stat.ethz.ch/pipermail/r-help/2007-February/125569.html) that the `data.frame` function will convert its arguments into `vector`, so we cannot use that but must rather build up the data frame by columns.

```{r}
df <- data.frame(dummy = 1:length(key8))
df$key8 <- key8
df$payload <- block.list
df$dummy <- NULL

knitr::kable(df)
```

We see that this works. Cool!

## Convenient function

We want to have a single function that just does the whole job. So here it is. Not sure whether you use Roxygen, but classic TeX-style R documentation can be generated from this using `roxygen2::roxygenise()`.

```{r}
#' Reads a Pandas HDF5 file from the contraction code.
#'
#' @param filename Path to HDF5 file
#' @param root Name of the root element to read from
read.pandas.hdf5 <- function (filename, root = 'data') {
    # Load the data and extract the relevant HDF5 entry.
    data <- h5read(filename, root)
    
    # Extract the number of time-slices.
    t <- length(data$axis0_level1)
    
    # Convert the payload into complex numbers.
    block <- data$block0_values$r + 1i * data$block0_values$i
    
    # Convert each row in the payload into a 2D matrix and then form a list
    # from all of the rows.
    block.list <- lapply(1:ncol(block), function (i) t(matrix(block[ , i], nrow = t)))
    
    # Find out how many axis1 elements there are.
    num.meta.cols <- sum(startsWith(names(data), 'axis1_label'))
    
    # We need to have a dummy variable to initialize the `data.frame` because
    # otherwise it would not have the correct length. Also the function
    # convertes, this would destroy our matrices and factors.
    df <- data.frame(dummy = 1:length(block.list))
    
    # Iterate through the meta data colums, extract them from the data
    # structure, “inflate” them and add them as `factor` objects to the
    # data frame.
    for (i in 0:(num.meta.cols - 1)) {
        level.name <- sprintf('axis1_level%d', i)
        label.name <- sprintf('axis1_label%d', i)
        col.name <- sprintf('C%d', i)
        
        level <- data[[level.name]]
        label <- data[[label.name]]
        
        df[col.name] <- factor(x = level[label + 1], levels = level)
    }
    
    # We want to have the payload at the end, so we just add it here.
    df$payload <- block.list
    
    # The dummy is not needed any more, we can delete it.
    df$dummy <- NULL
    
    return (df)
}
```

Try it out:

```{r}
test <- read.pandas.hdf5(filename)
test
```

# Sanity check

If the data was extracted correctly, plotting a single correlator should give a meaningful picture.

```{r}
one.corr <- test$payload[[1]]
m <- apply(one.corr, 2, mean)
plot(Re(m), log = 'y',
     main = 'Correlator for some quantum numbers',
     xlab = 'Time slice',
     ylab = 'Real part')
```

This looks sensible, so likely it is correct. Also the matrix dimensions are *configuration* and *time slice*, just as in R-hadron.